Computational methods in landscape ecology

### 2024-09-20 Sections/03-connectivity.qmd
Adding comments AdF

Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, quantifying connectivity has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
Generally, movement data from individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
Thus, in practice connectivity is primarily determined by indirect estimations.
@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal characteristics of species, and iii) actual connectivity related to observations of individuals moving through a landscape.

@@

Landscape metrics are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, yet functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].

@@

Many modern connectivity estimation approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational] and quantifying cost values for each cell using movement data.
This can included expert opinion [widely used due to low effort, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).

@@

Contrastingly, circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
In this approach, the landscape is represented as a network of electrical nodes connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
It can describes isolation by measuring multiple low- or high resistance pathways and identify areas of high movement probability highlighting important corridors and bottlenecks in the landscape [@Cushman2013BiologicalCorridors].

However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

@@

Following the formative work of @Urban2001LandscapeConnectivity, graph theory has developed into a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
Graph theory and connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
For example, they can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].

Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
The edges can be binary, representing whether a connection between two nodes exists or not, or weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment] which is based on a binary connection model [@Saura2010CommonCurrency].

Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
Furthermore, the modelling framework initially only applied to single species is increasingly being used to model multispecies landscapes to account for interspecific movement abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
In addition, modern multiple-layer graph approaches offer to model not only spatial, but spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and how these impacts affect biodiversity patterns [@Uroy2021HowAssess].

However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge due to various ways how to construct a graph and different data sources.
In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types which resulted in different distribution of connectivity values.

A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [Julia\; @vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [Julia\;@Landau2021OmniscapejlSoftware], *lconnect* [R\; @Mestre2023LconnectPackage], *grainscape* [R\; @Chubaty2020ToolboxGrainscape], *Circuitscape* [@Anantharaman2019CircuitscapeJulia], *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], among others.

>>>>

Although multiple approaches that measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], the technical and conceptual quantification of connectivity has proven challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
Generally, movement data from individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
Thus, in practice connectivity is primarily determined by indirect estimations.
@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal characteristics of species, and iii) actual connectivity related to observations of individuals moving through a landscape.

@@

Landscape metrics are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
Additionally, there has been a recent increase in focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, yet functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].

@@

Many modern connectivity estimation approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational] and quantifying cost values for each cell using movement data.
This can include expert opinion [widely used due to low effort, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).

@@

Contrastingly, circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
In this approach, the landscape is represented as a network of electrical nodes connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
It can describe isolation by measuring multiple low- or high resistance pathways and identify areas of high movement probability highlighting important corridors and bottlenecks in the landscape [@Cushman2013BiologicalCorridors].

However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments have shifted towards a combination of both resistance and processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

@@

Following the formative work of @Urban2001LandscapeConnectivity, graph theory has become a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
Graph theory and related connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
For example, graph theory based metrics can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].

Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
The edges can be binary, i.e., representing whether a connection between two nodes either exists or does not exists, or edges can be weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment] which is based on a binary connection model [@Saura2010CommonCurrency].

Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
Furthermore, the modelling framework was initially only applied to single species but is increasingly being used to model multispecies landscapes to account for interspecific movement abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
In addition, modern multiple-layer graph approaches enable the modelling of not only spatial, but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].

However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge from different methods used to construct a graph or from different data sources.
In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types which resulted in different distribution of connectivity values.

A large number of graph theory based software applications have been developed to estimate connectivity including *ConScape.jl* [Julia\; @vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [Julia\;@Landau2021OmniscapejlSoftware], *lconnect* [R\; @Mestre2023LconnectPackage], *grainscape* [R\; @Chubaty2020ToolboxGrainscape], *Circuitscape* [@Anantharaman2019CircuitscapeJulia], *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], among others.

### 2024-09-20 Sections/04-sampling.qmd
Adding comments AdF

Directly related to the location and extent of sampling sites is the issue of overlapping landscapes which can result in a lower range of landscape heterogeneity, lack of statistical independence, or pseudo-replications [@Hurlbert1984PseudoreplicationDesign; @Zuckerberg2020ReviewOverlapping].
Nonetheless, the lack of independence between sites is likely more related to spatial autocorrelation and thus not necessarily prevented by using non-overlapping landscapes [@Zuckerberg2020ReviewOverlapping].
Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures (e.g., Moran's I) between sampling approaches or spatial scales [@Pasher2013OptimizingLandscape].
Furthermore spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
If spatial autocorrelation is detected, researchers may either consider further data collection, or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].

>>>>

Directly related to the location and extent of sampling sites is the issue of overlapping landscapes which can result in a lower range of landscape heterogeneity, lack of statistical independence, or pseudo-replications [@Hurlbert1984PseudoreplicationDesign; @Zuckerberg2020ReviewOverlapping].
Nonetheless, the lack of independence between sites is likely more related to spatial autocorrelation and thus not necessarily prevented by using non-overlapping landscapes [@Zuckerberg2020ReviewOverlapping].
Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures (e.g., Moran's I) between sampling approaches or spatial scales [@Pasher2013OptimizingLandscape].
Furthermore spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
If spatial autocorrelation is detected, researchers may either consider further data collection, or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].

### 2024-09-20 Sections/05-simulations.qmd
Adding comments AdF

Simulation models are a powerful tool to study complex, adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
Due to the usual spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes], and are generally simpler than ecological simulations models [@Hess2020PioLaGPiosphere].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels], or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise], or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular], or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [e.g., @Day2020IndividualbasedModeling].
In order to facilitate future model developments, recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling], or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The Digital Twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins], and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
Thus, Digital Twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].

>>>>

Simulation models are a powerful tool to study complex, adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
Thus, simulations can be seen as experimental systems that allow all imaginable manipulations in order to advance theoretical developments or test hypotheses, which would be impossible in natural systems [@Peck2004SimulationExperiment].
Due to the spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general, simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Landscape simulators are typically used to generate null hypotheses, baselines, or scenarios that allow the users to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes], and are generally simpler than ecological simulation models [@Hess2020PioLaGPiosphere].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels], or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise], or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes, several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular], or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent landscape generators have allowed for the simulation of patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surrounding watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [e.g., @Day2020IndividualbasedModeling].
In order to facilitate future model developments, recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling], or reusable building blocks [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches that couple or even integrate different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

The digital twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins], and there is an increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
Thus, digital twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].

### 2024-09-20 Sections/06-genetics.qmd
Adding comments AdF

Genetic data can be applied to complement ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetic approaches have been used traditionally to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].

@@

Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], among others.
Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].

>>>>

Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].

@@

Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], *landgenreport* [@Gruber2015LandgenreportNew], *adegenet* [@Jombart2008AdegenetPackage], or *ape* [@Paradis2019Ape50], among others.
Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].
